Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-06-27 Paper.Rmd
replaced Gaussian by Normal and some text addition

<!--**SG: I think normal copula would be better as "N" is the abbreviation for that but for "G" means gumbel not gaussian.
If you agree with me, then I will made all "gaussian" as "normal" copula in both main and supmat.-->

@@

*Goes in the last paragraph of the intro:*
We address the following questions. 1) Do datasets in ecology and related fields have non-Gausian copula srtructure? Do they show tail dependence distinct from that of a Gaussian copula, and in particular do they show asymmetric tail dependence? 2) If so, what are some possible causes of non-Gaussian copula structure? And 3) what are the consequences of non-Gaussian copula structure and tail dependence for our understanding of biology and for applications?

@@

The first part or question 1 from the Introduction was answered for a bivariate dataset 
$(x_i,y_i)$ for $i=1,...,n$ by comparing the AIC weights of the non-Gaussian copula
families with that of the Gaussian family. The same was also done using BIC.
The second part of question 1 was answered by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of a Gaussian copula, both of which 
are 0. The third part of question 1
was answered by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for a Gaussian copula, again 0.

For multivariate datasets, e.g., the aphid abundance or first flight data, 
each of which consists of time series 
from 10 locations...
<!--***DAN: need to fill this in, prrobably enough to be brief, since these were just done 
pairwise-->

@@

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way. The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the algorithm had two versions, one for preserving each correlation coefficient), 
but had Gaussian copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were Gaussian, but the data were otherwise 
statistically unchanged.
Details of how surrogates were prroduced are in Appendix \ref{SM-surrog_test}, and code is in [`copsurrog2d`](https://github.com/sghosh89/BIVAN/blob/master/copsurrog2d.R).
<!--***DAN: I think the below material belongs in supp mat, because the reader can get the
main idea via the above text, and only some readers will be interested in the detailed 
algorithm. Please move it there, and then make the above links XYZ and ABC into live links
to the appropriate locations in the sup mat.

@@

For multivariate datasets, a spatial resampling scheme was used instead of the surrogate 
approach described above.
The scheme is identical to that proposed by @bjornstadfalck2001 for application in their now 
widely used spline correlogram method; the same scheme was also used by @walteretal2017 
as part of a method for 
performing model selection among matrix regression models. For each pair of 
locations $i \neq j$, the statistic
of interest, here denoted $s_{ij}$ ($\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, 
$\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, 
$\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$) was computed for those locations. The 
values $s_{ij}$ were then arranged in
a matrix, $S$, with diagonal entries NA. The mean of the non-NA entries of $S$ was computed. For 
$\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, or $\Dsq_{l_b,u_b}$ this mean characterized the 
average strength of 
dependence, across all pairs of distinct locations, for data in the portion of the distributions 
given by $l_b$ and $u_b$. For $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$, the mean characterized average
asymmetry of tail dependence across all pairs of locations. 
Confidence intervals for the values of $\mean(S)$ were then computed as quantiles of a 
distribution of bootstrapped values. If $\mean(S)$ was significantly different from 0, as
revealed by these confidence intervals, for the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, 
or $\Dsq_u-\Dsq_l$, then tail dependence was significantly asymmetric, in contrast to a 
Gaussian null hypothesis, and answering question 1 from the Introduction.
Bootstrapping was performed by resampling the locations of measurement, with replacement. 
This corresponds to 
resampling, with replacement, the rows and columns of $S$, and is described in detail elsewhere 
[@bjornstadfalck2001 ; @walteretal2017].

@@

- Do datasets in ecology and related fields have non-Gausian copula structure?
- Do they show tail dependence distinct from that of a Gaussian copula?
- In particular do they show asymmetric tail dependence?

@@

The test of indepedence gave $p=$ `r BivMS_res_Loecke$IndepTestRes` (to within the precision available 
from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_soilCNfit}). The table shows that several copulas are much better supported (higher AIC and BIC weights) than the Gaussian copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail dependence statistic was `r BivMS_res_Loecke$relLTdep_AICw` and the model-averaged upper-tail dependence statistic was `r BivMS_res_Loecke$relUTdep_AICw`. These values are distinct from what a Gaussian copula would give (i.e., 0), answering the second part of question 1 for these data. Furthermore, the difference of these two quantities (lower minus upper), `r BivMS_res_Loecke$relLTdep_AICw-BivMS_res_Loecke$relUTdep_AICw`, was different from 0, answering the third part of question 1.

Although these results do convincingly show non-Gaussian copula structure, we should take the tail dependence results with a grain of salt because even the lowest-AIC copula was a poor fit, giving $p$-value `r BivMS_res_Loecke$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_res_Loecke$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

@@

The test of indepedence gave $p=$ `r BivMS_BMR$IndepTestRes` (to within the precision available from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_BMRfit}). The table shows that BB6 and BB8 are much better supported (higher AIC and BIC weights) than the Gaussian copula, answering the first part of question 1 for these data.

@@

The test of indepedence gave $p=0$ <!--`r BivMS_contrasts_67$IndepTestRes` this is e-15 could not compile latex--> (to within the precision available 
from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_contrasts_67fit}). The table shows that several copulas (mostly J and SC) are much better supported (higher AIC and BIC weights) than the Gaussian copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail dependence statistic was `r BivMS_contrasts_67$relLTdep_AICw` and the model-averaged upper-tail dependence statistic was `r BivMS_contrasts_67$relUTdep_AICw`. These values are distinct from what a Gaussian copula would give (i.e., 0), answering the second part of question 1 for these data. Furthermore, the difference of these two quantities (lower minus upper), `r BivMS_contrasts_67$relLTdep_AICw-BivMS_contrasts_67$relUTdep_AICw`, was different from 0, answering the third part of question 1.

These results do convincingly show non-Gaussian copula structure, giving $p$-value `r BivMS_contrasts_67$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_contrasts_67$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

@@

<!--Question 2 analyses-->
## Some possible causes of non-Gaussian copula structure

@@

<!--Question 3 analyses-->
## Consequences of non-Gaussian copula structure and tail dependence for our understanding of biology and for applications?

@@

In order to study the effects of non-gaussian copula structure on extinction risk for population of a species at different locations with different dispersal rules, we examined the following generalized multiple patch model where local population dynamics at each patch were governed by the Lewton-Cohen model ($r=0$) or Ricker model ($r \neq 0$) as the case may be;
<!--did I write correct spelling for lewton model?-->

>>>>

<!--**SG: I think normal copula would be better as "N" is the abbreviation for that but for "G" means gumbel not gaussian.
I made all "Gaussian" as "Normal" copula in both main and supmat.-->

@@

*Goes in the last paragraph of the intro:*
We address the following questions. 1) Do datasets in ecology and related fields have non-Gausian copula srtructure? Do they show tail dependence distinct from that of a Normal copula, and in particular do they show asymmetric tail dependence? 2) If so, what are some possible causes of non-Normal copula structure? And 3) what are the consequences of non-Normal copula structure and tail dependence for our understanding of biology and for applications?

@@

The first part or question 1 from the Introduction was answered for a bivariate dataset 
$(x_i,y_i)$ for $i=1,...,n$ by comparing the AIC weights of the non-Normal copula
families with that of the Normal family. The same was also done using BIC.
The second part of question 1 was answered by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of a Normal copula, both of which 
are 0. The third part of question 1
was answered by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for a Normal copula, again 0.

For multivariate datasets, e.g., the aphid abundance or first flight data, 
each of which consists of time series from 10 locations, we carried out the same bivariate analysis
for all possible pairwise combinations of those locations and compared AIC-weights and tail-dependence of best-fit(minimum AIC) copula with that of the Normal copula. We also counted for how many location pairs model -averaged lower tail dependence is greater than model-averaged upper-tail dependence or vice-versa. For example, for green-spruce aphid abundance data we found major lower tail dependence whereas leaf-curling plum aphid's first flight data had significant upper tail dependence at all location pairs (for details see Table \ref{tab_multivar_summary}).
<!--***DAN: need to fill this in, prrobably enough to be brief, since these were just done 
pairwise-->
<!--**SG: done-->

@@

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way. The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the algorithm had two versions, one for preserving each correlation coefficient), 
but had Normal copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were Normal, but the data were otherwise 
statistically unchanged.
Details of how surrogates were prroduced are in Appendix \ref{SM-surrog_test}, and code is in [`copsurrog2d`](https://github.com/sghosh89/BIVAN/blob/master/copsurrog2d.R).
<!--***DAN: I think the below material belongs in supp mat, because the reader can get the
main idea via the above text, and only some readers will be interested in the detailed 
algorithm. Please move it there, and then make the above links XYZ and ABC into live links
to the appropriate locations in the sup mat.

@@

For multivariate datasets, a spatial resampling scheme was used instead of the surrogate 
approach described above.
The scheme is identical to that proposed by @bjornstadfalck2001 for application in their now 
widely used spline correlogram method; the same scheme was also used by @walteretal2017 
as part of a method for 
performing model selection among matrix regression models. For each pair of 
locations $i \neq j$, the statistic
of interest, here denoted $s_{ij}$ ($\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, 
$\Dsq_{l_b,u_b}$, $\cor_l-\cor_u$, 
$\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$) was computed for those locations. The 
values $s_{ij}$ were then arranged in
a matrix, $S$, with diagonal entries NA. The mean of the non-NA entries of $S$ was computed. For 
$\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, or $\Dsq_{l_b,u_b}$ this mean characterized the 
average strength of 
dependence, across all pairs of distinct locations, for data in the portion of the distributions 
given by $l_b$ and $u_b$. For $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, or $\Dsq_u-\Dsq_l$, the mean characterized average
asymmetry of tail dependence across all pairs of locations. 
Confidence intervals for the values of $\mean(S)$ were then computed as quantiles of a 
distribution of bootstrapped values. If $\mean(S)$ was significantly different from 0, as
revealed by these confidence intervals, for the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, 
or $\Dsq_u-\Dsq_l$, then tail dependence was significantly asymmetric, in contrast to a 
Normal null hypothesis, and answering question 1 from the Introduction.
Bootstrapping was performed by resampling the locations of measurement, with replacement. 
This corresponds to 
resampling, with replacement, the rows and columns of $S$, and is described in detail elsewhere 
[@bjornstadfalck2001 ; @walteretal2017].

@@

- Do datasets in ecology and related fields have non-Normal copula structure?
- Do they show tail dependence distinct from that of a Normal copula?
- In particular do they show asymmetric tail dependence?

@@

The test of indepedence gave $p=$ `r BivMS_res_Loecke$IndepTestRes` (to within the precision available 
from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_soilCNfit}). The table shows that several copulas are much better supported (higher AIC and BIC weights) than the Normal copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail dependence statistic was `r BivMS_res_Loecke$relLTdep_AICw` and the model-averaged upper-tail dependence statistic was `r BivMS_res_Loecke$relUTdep_AICw`. These values are distinct from what a Normal copula would give (i.e., 0), answering the second part of question 1 for these data. Furthermore, the difference of these two quantities (lower minus upper), `r BivMS_res_Loecke$relLTdep_AICw-BivMS_res_Loecke$relUTdep_AICw`, was different from 0, answering the third part of question 1.

Although these results do convincingly show non-Normal copula structure, we should take the tail dependence results with a grain of salt because even the lowest-AIC copula was a poor fit, giving $p$-value `r BivMS_res_Loecke$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_res_Loecke$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

@@

The test of indepedence gave $p=$ `r BivMS_BMR$IndepTestRes` (to within the precision available from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_BMRfit}). The table shows that BB6 and BB8 are much better supported (higher AIC and BIC weights) than the Normal copula, answering the first part of question 1 for these data.

@@

The test of indepedence gave $p=0$ <!--`r BivMS_contrasts_67$IndepTestRes` this is e-15 could not compile latex--> (to within the precision available 
from `BiCopIndTest` in the `VineCopula` package). We fitted several copulas
and got AIC values and upper- and lower-tail dependency estimates based on the fitted copulas (Table \ref{SM-tab_contrasts_67fit}). The table shows that several copulas (mostly J and SC) are much better supported (higher AIC and BIC weights) than the Normal copula, answering the first part of question 1 for these data.

The model-averaged (AIC weights used for model averaging) lower-tail dependence statistic was `r BivMS_contrasts_67$relLTdep_AICw` and the model-averaged upper-tail dependence statistic was `r BivMS_contrasts_67$relUTdep_AICw`. These values are distinct from what a Normal copula would give (i.e., 0), answering the second part of question 1 for these data. Furthermore, the difference of these two quantities (lower minus upper), `r BivMS_contrasts_67$relLTdep_AICw-BivMS_contrasts_67$relUTdep_AICw`, was different from 0, answering the third part of question 1.

These results do convincingly show non-Normal copula structure, giving $p$-value `r BivMS_contrasts_67$GofRes_CvM` (to within the precision available from `BiCopGofTest` in the `VineCopula` package) in the Cramer-von Mises-based test of the goodness of its fit, and $p$-value `r BivMS_contrasts_67$GofRes_KS` in the Kolmogorov-Smirnov-based test of the goodness of fit. The next section looks at tail dependence in a non-parametric way.

@@

<!--Question 2 analyses-->
## Some possible causes of non-Normal copula structure

@@

<!--Question 3 analyses-->
## Consequences of non-Normal copula structure and tail dependence for our understanding of biology and for applications?

@@

In order to study the effects of non-Normal copula structure on extinction risk for population of a species at different locations with different dispersal rules, we examined the following generalized multiple patch model where local population dynamics at each patch were governed by the Lewton-Cohen model ($r=0$) or Ricker model ($r \neq 0$) as the case may be;
<!--did I write correct spelling for lewton model?-->

### 2018-06-27 SupportingInformation.Rmd
replaced Gaussian by Normal and some text addition

<!--
# Question 1 analyses for each dataset
Recall that question 1 is: Do datasets in ecology and related fields have non-Gausian copula structure? Do they show tail dependence distinct from that of a Gaussian copula? In particular do they show asymmetric tail dependence?

@@

<!--
# Question 3 analyses for each dataset and theoretical analyses
Recall that question 3 is: What are the consequences of non-Gaussian copula structure and tail dependence for our understanding of biology and for applications?
## Skewness of the spatial mean
### Aphid count \label{skewness_count}
### Aphid first flight \label{skewness_ff}
### UK Sea plankton \label{skewness_pns}
### Methane-flux data \label{skewness_methane}
                   ----moved to Paper.Rmd----
-->

>>>>

<!--
# Question 1 analyses for each dataset
Recall that question 1 is: Do datasets in ecology and related fields have non-Gausian copula structure? Do they show tail dependence distinct from that of a Normal copula? In particular do they show asymmetric tail dependence?

@@

<!--
# Question 3 analyses for each dataset and theoretical analyses
Recall that question 3 is: What are the consequences of non-Normal copula structure and tail dependence for our understanding of biology and for applications?
## Skewness of the spatial mean
### Aphid count \label{skewness_count}
### Aphid first flight \label{skewness_ff}
### UK Sea plankton \label{skewness_pns}
### Methane-flux data \label{skewness_methane}
                   ----moved to Paper.Rmd----
-->
